On the Minimum Core Mass for Giant Planet Formation at Wide Separations

### 2014-02-12 tex/PaperI/ms.tex
text edits up to 5.2 some (not all) changes to response to match

To facilitate comparison with our numerical results, we rescale the analytic results as a rough correction for the neglect of self-gravity.  The goal of the rescaling is not to get a perfect agreement, but to aid in the interpretation of general trends.  To rescale, we simply assume that runaway occurs at a modified crossover mass, $M_{\rm atm} = f M\co$.
%AY: improve this discussion to make more clear that we are not actually implying that runaway growth begins early and make vernally less rambling.
\textbf{Here $f<1$, as growth is slower in our analytic model due to the neglect of self-gravity. For the purpose of this scaling, we are only interested in atmospheric growth up to the modified crossover mass defined above. We are thus not concerned about evolution to yet higher masses where the analytic approximation becomes inaccurate.} Mathematically, this choice of crossover mass amounts to replacing $\xi \rightarrow f\xi$ (cf. Equations \ref{eq:crit} and \ref{eq:Pcbc}).  We choose $f = 0.13$, which normalizes the analytic and numerical models at 10 AU, for standard parameters.  The revised scalings are
\begin{subeqnarray}
t_{\rm run} & \approx & 3 \times 10^6 {F_T^{5/2}  F_\kappa \left(\xi \over 3.4\right)^2  \over \mcn{10}^{5/3} \aun{10}^{15 / 14}} \yr  \slabel{eq:tcoolanf} \\
\MC & \approx & 8\, {F_T^{3/2} F_\kappa^{3/5}   \left(\xi \over 2.6 \right)^{6/5} \over \aun{10}^{9 / 14}} \; M_\oplus\ \, ,  \slabel{eq:Mcritf}
\end{subeqnarray} 
now using $t_{\rm run}$ instead of $t_{\rm co}$ for the runaway growth timescale, which is no longer tied to the crossover mass. \textbf{Here $t_{\rm run}$ is the time at which $M_{\rm atm}=f M\co$.} The numerical models are more accurate and self-consistent because they measure runaway growth directly, instead of relying on any assumed $M_{\rm atm}/M\co$ threshold.  When plotting the analytic scalings, we include the variations in $\xi$ from \Eq{eq:xidef}.

@@

\textbf{The atmospheric structure is well approximated by our non-self-gravitating, analytic solutions.  Deep in convective zones, \Eq{eq:Tconvdeep} gives $T \propto r^{-1}$ and $P\propto r^{-1/\delad}$.  This behavior is seen in all solutions in \Fig{fig:profiles} for $r\ll R\cb$. Near the core, where $m(r) \approx M\co$, \Eq{eq:deep} gives an expression for the core temperature, $T\co=\frac{G M\co}{C_{\rm P} R\co}$, which holds regardless of the atmosphere mass, as shown in Figure \ref{fig:profiles}. As the RCB is approached, \Eq{eq:Tconvdeep} is no longer valid, particularly for large envelope masses. Instead, these high mass solutions show a slightly flatter profile in $T$ and also in $P \propto T^{1/\delad}$, as explained by \Eq{eq:TPsg}.}

@@

At later stages of evolution, the numerical model in \Fig{fig:Ltplot} shows a flat luminosity with increasing mass and also time (not shown).  By contrast, the non-self-gravitating analytic model gives a luminosity that continues to drop as the atmosphere becomes more massive.  To understand this difference, consider the scaling of \Eq{eq:Lcb}, $L\cb \propto M\cb T\cb^4/ (\kappa\cb  P\cb)$, which holds in both cases.   Accounting for the higher enclosed mass in the self-gravitating model gives a somewhat higher luminosity, as desired.  However, the main effect is that \Eq{eq:crit} -- which describes a nearly linear relation between atmospheric mass and RCB pressure --  breaks down for self-gravitating solutions.  This behavior can be seen in the top panel of \Fig{fig:profiles} where the $P\cb$ increases significantly from 5.10 to 6.0 $M_\oplus$, but only increases relatively modestly with further growth to 8.99 $M_\oplus$.   In the higher mass solutions, the relatively low $P\cb$ values (and thus the relatively high luminosities) require an outward shift in $R\cb$, as shown in \Fig{fig:profiles}. \textbf{This shift does not occur in the analytic solution, where $R\cb$ continually decreases with mass (cf. Equations \ref{eq:crit} and \ref{eq:PcbRcb}).}

The accelerated growth in the numerical model, as shown in the bottom panel of \Fig{fig:Ltplot}, is also a direct result of the higher cooling luminosities in the self-gravitating models.  We see that self-gravity strongly affects atmospheric evolution, even well before the crossover mass, when $M_{\rm atm}=M\co$. \textbf{The early divergence between the analytic and numerical curves in Figure \ref{fig:Ltplot} shows that self-gravity becomes important even when the atmosphere is only a few percent as massive as the core. Our analytic model starts deviating from the numerical model at slightly larger $M_{\rm atm}/M\co$ closer to the star (not shown). However, we have found that self-gravity is always significant at $M_{\rm atm} \leq 10\% M\co$ in our regime of interest.}

@@

%We now use our structure and evolution models to calculate the necessary conditions for gas giant formation by core accretion.   
\textbf{We now use our structure and evolution models to estimate the timescales and core masses necessary for giant planet formation, for our assumed disk model, nebular gas properties and dust grain opacity.}
Adopting a median disk lifetime of 3 Myr, we examine the conditions for a fixed mass core to undergo runaway gas accretion within the lifetime of the gas disk.  Our results for atmosphere growth times -- for a range of core masses and disk conditions -- are presented in \S\ref{sec:tcross}.  Section \ref{sec:critcore} gives our results for critical core masses, the minimum masses needed to accrete a massive atmosphere within the disk lifetime.

@@

The effect of mean molecular weight is also shown in \Fig{fig:tvsMcomp}.  A lower $\mu$ gives longer growth times, because more cooling is required to compress the atmosphere.  This effect is explained by the increase in the atmospheric scale height $\propto 1/\mu$ while the Bondi radius decreases, $\RB \propto \mu$. \textbf{The larger atmospheric scale height due to a lower $\mu$ results in a larger pressure depth at the RCB, $P\cb$, and thus a lower $R\cb$, cf. \Eq{eq:PcbRcb}.} For $\mu = 2.0$, which represents the idealized case of an H$_2$ atmosphere completely devoid of Helium, $t_{\rm run}$ increases by factors of $\sim 2 - 3$.  Thus fairly drastic changes in atmospheric composition are required for $\mu$ to significantly affect core accretion timescales. \textbf{While we note that changes in the mean molecular weight of the gas also affect the adiabatic gradient $\delad$, we defer the discussion of the influence of $\delad$ on atmospheric evolution to Piso, Youdin \& Murray-Clay (in prep., hereafter Paper II).}

@@

\textbf{A larger disk lifetime further reduces $M_{\rm crit}$. Recent studies have shown that gas disks may live up to $\sim$$10-12$ Myr \citep{bell13}. As $M_{\rm crit} \sim t\di^{-3/5}$ (cf. Equation \ref{eq:tcoolan}), a disk lifetime of 10 Myrs decreases the critical core mass by a factor of two.}

@@

\section{Neglected Effects}\label{sec:neglected}

>>>>

To facilitate comparison with our numerical results, \textbf{we rescale the analytic results as follows.   Since growth times are too slow in the non-self-gravitating analytic model, we modify the runaway grown criterion to occur at an effective crossover mass $M_{\rm atm} = f M\co$ with $f < 1$.  Specifically we choose $f = 0.13$, because with this value the analytic model gives the same critical core mass at 10 AU as the numerical model, for our standard choices of other parameters.   This prescription does not mean that runaway growth physically occurs at low atmosphere masses, nor does it cause perfect agreement between the models.  Rather by artificially accelerating growth in the analytic model, we can more easily compare the parameter scalings of the two models.  Mathematically, the modified crossover mass amounts to replacing $\xi \rightarrow f\xi$ in our analysis.}  With $f = 0.13$, the revised scalings are
\begin{subeqnarray}
t_{\rm run} & \approx & 3 \times 10^6 {F_T^{5/2}  F_\kappa \left(\xi \over 3.4\right)^2  \over \mcn{10}^{5/3} \aun{10}^{15 / 14}} \yr  \slabel{eq:tcoolanf} \\
\MC & \approx & 8\, {F_T^{3/2} F_\kappa^{3/5}   \left(\xi \over 2.6 \right)^{6/5} \over \aun{10}^{9 / 14}} \; M_\oplus\ \, ,  \slabel{eq:Mcritf}
\end{subeqnarray} 
now using $t_{\rm run}$, defined as the time when $M_{\rm atm} = f M\co$,  instead of $t_{\rm co}$ for the runaway growth timescale.  When plotting these rescaled analytic results, we include the variations in $\xi$ from \Eq{eq:xidef}.

%To facilitate comparison with our numerical results, we rescale the analytic results as a rough correction for the neglect of self-gravity.  The goal of the rescaling is not to get a perfect agreement, but to aid in the interpretation of general trends.  To rescale, we simply assume that runaway occurs at a modified crossover mass, $M_{\rm atm} = f M\co$.
%\textbf{Here $f<1$, as growth is slower in our analytic model due to the neglect of self-gravity. For the purpose of this scaling, we are only interested in atmospheric growth up to the modified crossover mass defined above. We are thus not concerned about evolution to yet higher masses where the analytic approximation becomes inaccurate.} Mathematically, this choice of crossover mass amounts to replacing $\xi \rightarrow f\xi$ (cf. Equations \ref{eq:crit} and \ref{eq:Pcbc}).  We choose $f = 0.13$, which normalizes the analytic and numerical models at 10 AU, for standard parameters.  The revised scalings are
%\begin{subeqnarray}
%t_{\rm run} & \approx & 3 \times 10^6 {F_T^{5/2}  F_\kappa \left(\xi \over 3.4\right)^2  \over \mcn{10}^{5/3} \aun{10}^{15 / 14}} \yr  \slabel{eq:tcoolanf} \\
%\MC & \approx & 8\, {F_T^{3/2} F_\kappa^{3/5}   \left(\xi \over 2.6 \right)^{6/5} \over \aun{10}^{9 / 14}} \; M_\oplus\ \, ,  \slabel{eq:Mcritf}
%\end{subeqnarray} 
%now using $t_{\rm run}$ instead of $t_{\rm co}$ for the runaway growth timescale, which is no longer tied to the crossover mass. \textbf{Here $t_{\rm run}$ is the time at which $M_{\rm atm}=f M\co$.} The numerical models are more accurate and self-consistent because they measure runaway growth directly, instead of relying on any assumed $M_{\rm atm}/M\co$ threshold.  When plotting the analytic scalings, we include the variations in $\xi$ from \Eq{eq:xidef}.

@@

\textbf{The atmospheric structure is well approximated by our non-self-gravitating, analytic solutions.  Deep in convective zones, thermal energy is a fixed fraction of the gravitational potential energy, giving $T \propto r^{-1}$ and $P\propto r^{-1/\delad}$  as in \Eq{eq:deep}.  This behavior is seen in \Fig{fig:profiles} for $r\ll R\cb$. Near the core, \Eq{eq:Tconvdeep} gives the core temperature, $T\co= G M\co / (C_{\rm P} R\co)$, unaffected by  the overlying atmosphere mass which does not contribute to the gravitational potential.  Closer to the RCB, self-gravity is no longer negligible, particularly for large envelope masses. Instead, these high mass solutions show a slightly flatter profile in $T$ and also in $P \propto T^{1/\delad}$, as explained by \Eq{eq:TPsg}.}

@@

At later stages of evolution, the numerical model in \Fig{fig:Ltplot} shows a flat luminosity with increasing mass and also time (not shown).  By contrast, the non-self-gravitating analytic model gives a luminosity that continues to drop as the atmosphere becomes more massive.  To understand this difference, consider the scaling of \Eq{eq:Lcb}, $L\cb \propto M\cb T\cb^4/ (\kappa\cb  P\cb)$, which holds in both cases.   Accounting for the higher enclosed mass in the self-gravitating model gives a somewhat higher luminosity, as desired.  However, the main effect is that \Eq{eq:crit} -- which describes a nearly linear relation between atmospheric mass and RCB pressure --  breaks down for self-gravitating solutions.  This behavior can be seen in the top panel of \Fig{fig:profiles} where the $P\cb$ increases significantly from 5.10 to 6.0 $M_\oplus$, but only increases relatively modestly with further growth to 8.99 $M_\oplus$.   In the higher mass solutions, the relatively low $P\cb$ values (and thus the relatively high luminosities) require an outward shift in $R\cb$, as shown in \Fig{fig:profiles}. \textbf{This shift does not occur in the analytic solution, where $R\cb$ continually decreases with atmospheric mass (cf. Equations \ref{eq:PcbRcb} and \ref{eq:crit}).}

The accelerated growth in the numerical model, as shown in the bottom panel of \Fig{fig:Ltplot}, is also a direct result of the higher cooling luminosities with self-gravity included.  \textbf{Even when $M_{\rm atm} / M\co$ is only few percent, i.e.\ well before the crossover mass, the effect of self-gravity is  quite evident.  Closer to the star, the effects of self-gravity are not as strong for low atmosphere masses.  Nevertheless our models all show that self-gravity noticeably accelerates growth for $M_{\rm atm} > 0.1 M\co$.}

@@

%We now use our structure and evolution models to calculate the necessary conditions for gas giant formation by core accretion.   
\textbf{We now use our structure and evolution models to estimate the timescales and minimum core masses for giant planet formation for a range of disk conditions and other model parameters.  Our results for atmosphere growth times -- the time for a core of fixed mass to undergo runaway gas accretion -- are presented in \S\ref{sec:tcross}.  Section \ref{sec:critcore} gives our results for critical core masses, the minimum values that trigger runaway atmospheric growth within a plausible disk lifetime, here 3 Myr.}

@@

The effect of mean molecular weight is also shown in \Fig{fig:tvsMcomp}.  A lower $\mu$ gives longer growth times, because more cooling is required to compress the atmosphere.  \textbf{This effect is intuitive since the atmospheric scale height $\propto 1/\mu$ is more extended for lower $\mu$.   The Bondi radius decreases as $\RB \propto \mu$ giving a smaller gravitational sphere of influence (and a weaker compression at $\RH$ when that is the more relevant scale).} For $\mu = 2.0$, which represents the idealized case of an H$_2$ atmosphere completely devoid of Helium, $t_{\rm run}$ increases by factors of $\sim 2 - 3$.  Thus fairly drastic changes in atmospheric composition are required for $\mu$ to significantly affect core accretion timescales. \textbf{In principle, changes in the mean molecular weight of the gas also affect the EOS, including $\delad$, but such effects are not considered here, see \S\ref{sec:EOS}.}

@@

\textbf{A larger disk lifetime further reduces $M_{\rm crit}$. Recent studies have shown that gas disks may live up to $\propto$$10-12$ Myr \citep{bell13}. As $M_{\rm crit} \sim t\di^{-3/5}$ (cf. Equation \ref{eq:tcoolan}), a disk lifetime of 10 Myrs decreases the critical core mass by a factor of two.  Of course we also require some fraction of the disk lifetime for the core to grow and possibly migrate.}

@@

\section{Neglected Effects}\label{sec:neglected}
\textbf{Since one goal of this paper was to obtain a detailed understanding of atmospheric evolution with a simple model, we have necessarily ignored many  effects of potential significance.  We briefly address the most important of these and note that a followup work, Piso, Youdin \& Murray-Clay (in prep., hereafter Paper II), will extend these models to address some of them.}
